---
title: "AdvancedR-functional operators"
author: "Luke O'Donnell"
date: "February 5, 2017"
output: html_document
---

# 12.1: Behavioural FO's

# Exercises  
  
1. Write a FO that logs a time stamp and message to a file every time a function is run.
```{r}
## The function will create a file in your current WD
logger <- function(f, filename, message) {
   force(f)
   filename_tmp <- paste(filename, basename(tempfile()), sep = "_")
   write(paste("created at:", Sys.time()), filename_tmp, append = TRUE)
   function(..., message = "Default mesage to each call") {
      write(paste0("used at: ", Sys.time(), ", ", message), filename_tmp, append = TRUE)
   }
}

mean2 <- logger(mean, "mean_log")
mean2(1:4, message = "first invocation")
mean2(5:8, "second invocation")

```
2. What does the following function do? What would be a good name for it?
```{r}
f <- function(g) {
   force(f)
   result <- NULL
   function(...) {
      if(is.null(result)) {
         result <<- g(...)
      }
      result
   }
}
runif2 <- f(runif)
runif2(10)
runif2(5)

```
This function stores & will always return the result of the first invocation of function ```g```, no matter what inputs are supplied to subsequent invocations of function ```g```. A good name for the function would be ```initial_call```  
  
3. Modify ```delay_by``` so that instead of delaying by a fixed amount of time, it ensures that a certain amount of time has elapsed since the function was last called. That is, if you called ```g <- delay_by(1, f); g(); Sys.sleep(2); g() there shouldn't be an extra delay.  
  
  
```{r}
delay_by <- function(elapse, f) {
   force(f)
   prev_time <- Sys.time()
}
```
